3.9: Logarithmic Regression

### 2019-11-21 static/slides/21-slides.Rmd
Fix errors on slides 21

- The .red[exponential function], $e^x$ or **$exp(x)$**, where base $e=2.71828...$

- .blue[Natural logarithm] is the inverse, $y=ln(x)$

@@

# The Natural Logarithm: Review

- **Exponents** are defined as
$$b^n=\underbrace{b \times b \times ... \times b}_{n}$$
where base $b$ is multiplied by itself $n$ times

@@

- **Logarithms** the inverse, defined as the exponents in the expressions above
$$\text{If } b^n=y\text{, then }log_b(y)=n$$
$n$ is the number you must raise $b$ to in order to get $y$

@@

@@

- Natural logs have a lot of useful properties:
  - $ln(\frac{1}{x})=-ln(x)$
  - $ln(ab)=ln(a)+ln(b)$
  - $ln(\frac{x}{a})=ln(x)-ln(a)$
  - $ln(x^a)=a \, ln(x)$
  - $\frac{d  \, ln \, x}{d \, x} = \frac{1}{x}$

@@

- An .shout[elasticity] between two variables, $E_{Y,X}$ describes the .whisper[responsiveness] of one variable to a change in another.

@@

$$\begin{align*}
E_{y,x}&=\frac{\% \Delta y}{\% \Delta x}\\
&=\cfrac{\frac{\Delta y}{y}}{\frac{\Delta x}{x}}\\
\end{align*}$$

@@

```{r}
gapminder<-gapminder %>%
  mutate(loggdp=log(gdpPercap))

gapminder
```

@@

- .shout[Linear-log model] has an independent variable ($X$) that is logged

@@

- $\hat{\beta_1}$: **a 1\% change in $X$** $\rightarrow$ **$\frac{\beta_1}{100}$ unit change in $Y$**

@@

- $\hat{\beta_1}$: a **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy

@@

- $\hat{\beta_1}$: a **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy

- A **25% fall in GDP** $\rightarrow$ a $(25 \times 0.0941)=$ **2.353 year decrease** in Life Expectancy

@@

$$\begin{align*}
\mathbf{Y}&=\beta_0+\beta_1 X\\
\beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{X}\\
\end{align*}$$

@@

- $\hat{\beta_1}$: **a 1 unit change in $X$** $\rightarrow$ **$\beta_1 \times 100\%$ change in $Y$**

@@

```{r}
gapminder <- gapminder %>%
  mutate(gdp_t=gdpPercap/1000,
         loglife=log(lifeExp))
```

@@

- $\hat{\beta_1}$: a **\$1 (thousand) change in (thousands of) GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3\% increase** in Life Expectancy

@@

- $\hat{\beta_1}$: a **\$1 (thousand) change in (thousands of) GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3\% increase** in Life Expectancy

- A **\$25K fall in GDP** $\rightarrow$ a $(25 \times 1.3\%)=$ **32.5\% decrease** in Life Expectancy

@@

- $\hat{\beta_1}$: a **\$1 (thousand) change in (thousands of) GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3\% increase** in Life Expectancy

- A **\$25K fall in GDP** $\rightarrow$ a $(25 \times 1.3\%)=$ **32.5\% decrease** in Life Expectancy

- A **\$100K rise in GDP** $\rightarrow$ a $(100 \times 1.3\%)=$ **130\% increase** in Life Expectancy

@@

$$\begin{align*}
\mathbf{Y}&=\beta_0+\beta_1 \mathbf{X}\\
\beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{\big(\frac{\Delta X}{X}\big)}\\
\end{align*}$$

@@

- $\hat{\beta_1}$: **a 1% change in $X$** $\rightarrow$ **$\beta_1\%$ change in $Y$**

@@

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$

@@

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$

- $\hat{\beta_1}$: a **$1\% change in GDP** $\rightarrow$ a **0.147\% increase** in Life Expectancy

@@

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$

- $\hat{\beta_1}$: a **$1\% change in GDP** $\rightarrow$ a **0.147\% increase** in Life Expectancy

- A **25\% fall in GDP** $\rightarrow$ a $(25 \times 0.147\%)=$ **3.675\% decrease** in Life Expectancy

@@

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.0147 \, \text{ln(GDP)}_i$$

- $\hat{\beta_1}$: a **$1\% change in GDP** $\rightarrow$ a **0.147\% increase** in Life Expectancy

- A **25\% fall in GDP** $\rightarrow$ a $(25 \times 0.147\%)=$ **3.675\% decrease** in Life Expectancy

- A **100\% rise in GDP** $\rightarrow$ a $(100 \times 0.147\%)=$ **14.7\% increase** in Life Expectancy

@@

| Variable | Mean | Std. Dev. |
|----------|------|-----------|
| Salary | $2,024,616 | $2,764,512 |
| Batting Average | 0.267 | $0.031 |
| Home Runs | 12.11 | 10.31 |

@@

- **Marginal effect** on $Y$ (in *standard deviations* of $Y$) from 1 *standard deviation* change in $X$
  - e.g. for 1 standard deviation increase in Batting Average, Salary increases by 0.14 standard deviations
  - $0.14\times 2,764,512=\$387,032$

@@

```{r}
gapminder<-gapminder %>%
  mutate(std_life = scale(lifeExp),
         std_gdp = scale(gdpPercap))

@@

- The .shout[F-statistic} is the test-statistic used to test joint hypotheses about regression coefficients with an .shout[F-test}

@@

- $F$ is an .whisper[analysis of variance (ANOVA)]
  - essentially tests whether $R^2$ increases statistically significantly as we go from the restricted model$\rightarrow$unrestricted model

@@

- $R^_r$: the $R^2$ from the **restricted model** (null hypothesis)
]

@@

- $R^_r$: the $R^2$ from the **restricted model** (null hypothesis)

@@

- .whisper[Key takeaway]: The bigger the difference between $(R^2_u-R^2_r)$, the greater the improvement in fit by adding variables, the larger the $F$!

@@

.font50[
```{r, echo=F, results="asis"}
summary(unrestricted_reg)
```
]

@@

- "Statistic" is the All F-test, "p-value" next to it is the p value from the F test!

>>>>

- The .red[exponential function], $Y=e^X$ or $Y=exp(X)$, where base $e=2.71828...$

- .blue[Natural logarithm] is the inverse, $Y=ln(X)$

@@

# The Natural Logarithm: Review I

- **Exponents** are defined as
$$b^n=\underbrace{b \times b \times \cdots \times b}_{n \text{ times}}$$
  - where base $b$ is multiplied by itself $n$ times

@@

- **Logarithms** are the inverse, defined as the exponents in the expressions above
$$\text{If } b^n=y\text{, then }log_b(y)=n$$
  - $n$ is the number you must raise $b$ to in order to get $y$

@@

# The Natural Logarithm: Review II

@@

- Natural logs have a lot of useful properties:

1. $ln(\frac{1}{x})=-ln(x)$

2. $ln(ab)=ln(a)+ln(b)$

3.  $ln(\frac{x}{a})=ln(x)-ln(a)$

4. $ln(x^a)=a \, ln(x)$

5.  $\frac{d  \, ln \, x}{d \, x} = \frac{1}{x}$

@@

- An .shout[elasticity] between two variables, $E_{Y,X}$ describes the .onfire[responsiveness] of one variable to a change in another

@@

$$E_{Y,X}=\frac{\% \Delta Y}{\% \Delta X} =\cfrac{\left(\frac{\Delta Y}{Y}\right)}{\left( \frac{\Delta X}{X}\right)}$$

@@

```{r, echo=T}
gapminder<-gapminder %>%
  mutate(loggdp=log(gdpPercap)) # log GDP per capita

gapminder %>% head() # look at it
```

@@

- .shout[Linear-log model] has an independent variable $(X)$ that is logged

@@

- .onfire[**Marginal effect of** `\\(\mathbf{X \rightarrow Y}\\)`: a **1%** change in `\\(X \rightarrow\\)` a `\\(\frac{\beta_1}{100}\\)` **unit** change in `\\(Y\\)`]

@@

- A **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy

@@

- A **1% change in GDP** $\rightarrow$ a $\frac{9.41}{100}=$ **0.0941 year increase** in Life Expectancy

- A **25% fall in GDP** $\rightarrow$ a $(-25 \times 0.0941)=$ **2.353 year decrease** in Life Expectancy

@@

$$\begin{align*}
\mathbf{ln(Y)}&=\beta_0+\beta_1 X\\
\beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{\Delta X}\\
\end{align*}$$

@@

- .onfire[**Marginal effect of** `\\(\mathbf{X \rightarrow Y}\\)`: a **1 unit** change in `\\(X \rightarrow\\)` a `\\(\beta_1 \times 100\\)` **%** change in `\\(Y\\)`]

@@

```{r, echo=T}
gapminder <- gapminder %>%
  mutate(gdp_t=gdpPercap/1000, # first make GDP/capita in $1000s
         loglife=log(lifeExp)) # take the log of LifeExp
gapminder %>% head() # look at it
```

@@

- A **$1 (thousand) change in GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3% increase** in Life Expectancy

@@

- A **$1 (thousand) change in GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3% increase** in Life Expectancy

- A **$25 (thousand) fall in GDP** $\rightarrow$ a $(-25 \times 1.3\%)=$ **32.5% decrease** in Life Expectancy

@@

- A **$1 (thousand) change in GDP** $\rightarrow$ a $0.013 \times 100\%=$ **1.3% increase** in Life Expectancy

- A **$25K fall in GDP** $\rightarrow$ a $(-25 \times 1.3\%)=$ **32.5% decrease** in Life Expectancy

- A **$100K rise in GDP** $\rightarrow$ a $(100 \times 1.3\%)=$ **130% increase** in Life Expectancy

@@

$$\begin{align*}
\mathbf{ln(Y)}&=\beta_0+\beta_1 \mathbf{ln(X)}\\
\beta_1&=\cfrac{\big(\frac{\Delta Y}{Y}\big)}{\big(\frac{\Delta X}{X}\big)}\\
\end{align*}$$

@@

- .onfire[**Marginal effect of** `\\(\mathbf{X \rightarrow Y}\\)`: a **1%** change in `\\(X \rightarrow\\)` a `\\(\beta_1\\)` **%** change in `\\(Y\\)`]

- $\beta_1$ is the **elasticity** of $Y$ with respect to $X$!

@@

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$

@@

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$

- A **$1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy

@@

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$

- A **$1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy

- A **25% fall in GDP** $\rightarrow$ a $(-25 \times 0.147\%)=$ **3.675% decrease** in Life Expectancy

@@

$$\widehat{\text{ln(Life Expectancy)}}_i=2.864+0.147 \, \text{ln(GDP)}_i$$

- A **$1% change in GDP** $\rightarrow$ a **0.147% increase** in Life Expectancy

- A **25% fall in GDP** $\rightarrow$ a $(-25 \times 0.147\%)=$ **3.675% decrease** in Life Expectancy

- A **100% rise in GDP** $\rightarrow$ a $(100 \times 0.147\%)=$ **14.7% increase** in Life Expectancy

@@

| Variable | Mean | Std. Dev. |
|----------|------|-----------|
| Salary | $2,024,616 | $2,764,512 |
| Batting Average | 0.267 | 0.031 |
| Home Runs | 12.11 | 10.31 |

@@

- .onfire[Marginal effect] on $Y$ (in *standard deviations* of $Y$) from 1 *standard deviation* change in $X$
  - $\hat{\beta_1}$: a 1 standard deviation increase in Batting Average increases Salary by 0.14 standard deviations
    - $0.14 \times 2,764,512=\$387,032$
  - $\hat{\beta_2}$: a 1 standard deviation increase in Home Runs increases Salary by 0.48 standard deviations
    - $0.48 \times 2,764,512=\$1,326,966$

@@

```{r, echo=T}
gapminder<-gapminder %>%
  mutate(std_life = scale(lifeExp),
         std_gdp = scale(gdpPercap))

@@

- The .shout[F-statistic] is the test-statistic used to test joint hypotheses about regression coefficients with an .shout[F-test]

@@

- $F$ is an .onfire[analysis of variance (ANOVA)]
  - essentially tests whether $R^2$ increases statistically significantly as we go from the restricted model$\rightarrow$unrestricted model

@@

- $R^2_r$: the $R^2$ from the **restricted model** (null hypothesis)
]

@@

- $R^2_r$: the $R^2$ from the **restricted model** (null hypothesis)

@@

- .onfire[Key takeaway]: The bigger the difference between $(R^2_u-R^2_r)$, the greater the improvement in fit by adding variables, the larger the $F$!

@@

.font50[
```{r, echo=F}
summary(unrestricted_reg)
```
]

@@

- "**statistic**" is the All F-test, "**p.value**" next to it is the p value from the F test!
